A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-04 doc/popl2013/langdef.tex
reworded soundness theorem statement, changes to nondeterminism in operational semantics

A number of the rules are worth discussing in detail.

Because we would like to keep the heap typing $H$ constant, the semantics for the expression
new $T@r$ allow the choice of any memory location that falls within the right physical region
and has the appropriate heap type.  The soundness of the type system does not depend on that
location being one that isn't already in use, although most implementations will want to 
add that additional guarantee.

A similar choice appears in the semantics of the read operation.  If the location being read
from is within the clobber set $C$, the semantics appear to allow any value to be returned, as
long as it has the right type.  In reality, the choices are more constrained.  If it's not the
most recently written/reduced value from the current expression evaluation, it must at least be
a value written by some concurrent evaluation.  The constraint is applied further up in the 
expression tree when the result of the function is interleaved with other memory traces.

@@

\rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\

@@

\begin{thm}
For any expression $e$, if:
\begin{center}
$
\begin{array}{l} \\
\typeenv{e}{T} \\
\opsenv{e}{v}{E} \\
M \sim \Omega \\
L \sim_H \Gamma \\
S \sim H
\end{array}
$
\end{center}
then:
\begin{center}
$
\begin{array}{l}
v \sim_H M \llbracket T \rrbracket \\
E \sim H \\
M \vdash E : \Phi
\end{array}
$
\end{center}
\end{thm}

Although there are 23 different cases to be proven, many of them are similar, and benefit from
the use of the following lemmas:

@@

The complete soundness proof can be found here\cite{xyz} - for space reasons we just discuss
the general strategy here.

@@

\item Only three expressions have subexpressions that modify $M$ or $\Omega$.  For the
{\em partition} expression, the assumed consistency of the coloring can be shown to preserve
the region mapping consistency with respect to the constraints.  Similarly, the consistency of 
a region relation instance guarantees the maintenance of consistency in an {\em unpack}
expression.  Finally, the body of a called function executes with a different region mapping,
but with an initially-empty set of constraints, so they are trivially satisfied.

>>>>

There are a number of rules that appear to be nondeterministic, allowing multiple choices for
valid evaluation.  In reality, only the memory allocation case (i.e. new $T@r$) has a truly
nondeterministic choice (i.e. which currently-unused memory location of the right type to 
use).  A particular memory allocation algorithm is likely to be deterministic, but the
design of such algorithms is unrelated to the subject at hand, and by abstracting the location 
choice, we show that our results hold for any correct memory allocation algorithm.  (In fact,
our criteria are even weaker than that.  Our soundness result even holds if the location returned
by ``new $T@r$'' is one that is already in use, as long as the heap typing of that location is
correct.)

The remaining cases of apparent nondeterminism are the result of how we handle the semantics of
concurrent execution.  An operational semantics that explicitly tracks the state of all 
concurrently executing operations would allow precise specification of the result of each 
evaluation step, but would be extremely cumbersome.  Instead, we identify the cases in which
nonlocal effects (i.e. those that result from the concurrent evaluation of other expressions)
can impact the local evaluation, and make a ``guess'' of what value would result from an 
operation.  The check that the guess was consistent with the state of the system is deferred until
the previously nonlocal effects become local (i.e. contained within the evaluation of an enclosing
expression tree).  If the guess is not consistent, then the attempted proof of the subexpression's
behavior is flawed and must be reattempted.

In the case of ``read'' operations whose locations are in the clobber set $C$, we allow the local
evaluation to guess any value as long as it is of the right type.  The consistency of the value
is eventually checked by the $valid\_interleave$ predicate of some enclosing expression, but it
may not be the closest enclosing one.  The $valid\_interleave$ test itself allows some (apparent)
freedom of choice, deferring some of the consistency checks to the corresponding check further up
the expression tree.  The final case of apparent nondeterminism is in the choice of what 
additional locations to include in each subregion that results from a {\em partition} operation.
The expansion of the coloring set is necessary for the success of subsequent memory allocation
within a subregion, but that allocation might not occur within the {\em partition} expression's
body.  Although the correctness of these partitioning guesses cannot be fully verified until the
entire execution is known, when an inconsistency is encountered (i.e. there isn't a location of
the right type in the right region for a ``new $T@r$'' expression) it is possible to ``patch'' the 
proof (by adding a previously unnamed location to the appropriate expanded colorings and giving
it the right type in $H$) rather than having to start the proof over from the beginning.

@@

\opsenv{e_1}{K}{E_1} \\
K' = K \cup \{ (l_1, v_1), \ldots, (l_p, v_p) \}, \text{ where } \forall i \in [1,p]. (\forall v'. (l_i, v') \not\in K) \wedge (\forall j \in [1,p]. l_i \not= l_j) \wedge v_i \in [1,k] \\
\rho_i = \{ l : (l, i) \in K' \}, \text{ for } 1 \leq i \leq k \\

@@

\begin{thm}
Let $e$ be a well-typed Legion expression.  Let $M$ be a region mapping, $L$ be local value mapping,
$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
$$
\begin{array}{l}
M \sim \Omega \\
L \sim_H \Gamma \\
S \sim H
\end{array}
$$
If $v$ and $E$ are the value and memory trace resulting from a permissable terminating evaluation of
$e$, then:
$$
\begin{array}{l}
v \sim_H M \llbracket T \rrbracket \\
E \sim H \\
M \vdash E : \Phi
\end{array}
$$
\end{thm}

Due to space constraints, we limit the discussion to the general strategy of the proof.  The full
proof can be found here\cite{xyz}.  If $v$ and $E$ are the result of a permissible evaluation of
$e$, then a proof exists of:
$$\opsenv{e}{v}{E}$$
The soundness of the result is shown by induction over the structure of this proof.  For each of 
the 23 different forms a Legion expression can have, we assume the soundness of the subexpressions'
evaluations (provided we can show the consistency of the mappings and store used by that
subexpression) and prove that the result of the expression's evaluation is sound.  Although there
are a multitude of cases to be proven, many of them are similar, and benefit from
the use of the following lemmas:

@@

The proofs of these lemmas can also be found here\cite{xyz}.  The proof of the theorem is organized
into six pieces, one for each of the three consistency properties needed for subexpressions and one
for each of the three soundness properties we wish to show.

@@

\item Only three expressions have subexpressions that modify $M$ or $\Omega$ - those that don't
trivially satisfy the region mapping consistency requirement.  For the
{\em partition} expression, the assumed consistency of the coloring can be shown to preserve
the region mapping consistency with respect to the constraints.  Similarly, the consistency of 
a region relation instance guarantees the maintenance of consistency in an {\em unpack}
expression.  Finally, the body of a called function executes with a different region mapping,
but with an initially-empty set of constraints, so they are trivially satisfied.
